Fair Queuing Aware Congestion Control

### 2022-09-28 tex/paper.tex
Proofread

\begin{abstract}
Fair queuing is becoming increasingly prevalent in the internet and has been shown to improve performance in many circumstances. 
Performance could be improved even more if endpoints could detect the presence of fair queuing on a certain path and adjust their congestion control accordingly. 
If fair queuing is detected, the congestion control would not have to take cross traffic into account, which allows for more flexibility. 
In this paper, we develop the first algorithm that continuously checks if fair queuing is present on a path, with very high accuracy. 
When fair queuing is detected, a different congestion control can be chosen, which can result in reduced latency. In case fair queuing is detected, each flow can specify how much queuing delay it allows, meaning that each flow can choose its own tradeoff between throughput and latency. 
\end{abstract}

@@

In our previous work we proposed a technique which determines the presence of fair queuing at flow startup, which works as follows: 
If fair queuing is successfully detected at flow startup, a congestion control was used which aimed to keep queuing delay low (delay-based congestion control). 
While this delay-based congestion control achieved high throughput and low delay, similar to the Vegas congestion control algorithm \cite{brakmo_tcp_1995}, 
it was vulnerable to be outcompeted by other network flows sending more aggressively, such as \cite{cardwell_bbr_2016,dong_pcc_2015,ha_cubic_2008}.
This means that our delay-based congestion control performed well but only when it wouldn't have to compete with other flows. Thus it is only used if fair queuing is detected. 
If the absence of fair queuing is detected, our previous approach uses a more aggressive congestion control (specifically PCC \cite{dong_pcc_2015}), which can compete better with other, more aggressive network flows,
but doesn't keep delay as low as our delay-based congestion control.

@@

We would thus like to have an approach which 
\begin{itemize}
    \item \textbf{continuously checks} for the presence or absence of fair queuing, not only at flow startup. 
    \item \textbf{doesn't cause packet loss} while trying to determine if there is fair queuing or not. 
    \item can be \textbf{transparently used} on top of any congestion control algorithm. 
    \item lets each application \textbf{choose how much delay} it allows in case fair queuing is detected. Allowing a higher delay can result in higher throughput. 
\end{itemize}

@@

On the other side, if there is no fair queuing, when the sending rate exceeds the capacity of the bottleneck link, a queuing is going to build up and queuing delay is going to rise for \textbf{both} the dominant and the non-dominant subflow (\autoref{fig:illustration_no_fq}). 
In this case, Tonopah doesn't behave different from a network flow which doesn't have fair queuing detection. 
This can be a potential advantage which a Tonopah-enabled congestion control can have over other congestion control algorithms such as BBR, which has been shown to be unfair to other flows sometimes \cite{ware_modeling_2019,hock_experimental_2017}.

@@

Both subflows are controlled by the same congestion window. The differences in sending rates are achieved by using pacing. 
The basis of the implementation of Tonopah is \textit{Picoquic}\footnote{\url{https://github.com/private-octopus/picoquic/tree/master/picoquic}}. 
Picoquic includes support for Multipath QUIC and pacing, which we need for the implementation of Tonopah. Also, it implements congestion control algorithms. 
We base Tonopah on the NewReno implementation of Picquic. When fair queuing is detected, the congestion window is reduced by $\frac{1}{8}$. 
This worked well for our proof-of-concept implementation but one could also specify any other behavior in case fair queuing is detected or even do nothing at all.

@@

We emulate the network using \textit{mininet} and use Linux's \textit{pfifo} queuing discipline in the case there is no fair queuing and \textit{fq} for fair queuing. 
For evaluation we choose a simple network in which the sender is connected to the receiver via a switch.

@@

To evaluate Tonopah's performance under crosstraffic, we use iperf with the NewReno congestion control. We start iperf 4\,s before Tonopah so that it has time to saturate the link. 
\autoref{fig:no_fq_crosstraffic} and \autoref{fig:fq_crosstraffic} show that also in the presence of cross-traffic Tonpah's detection accuracy is virtually unchanged or even slightly better. We tried to use fq\_codel as in \autoref{fig:fq} but we encountered problems when using fq\_codel in combination with iperf3. Frequently, iperf3 would lose all packets, have a timeout after a couple of seconds and stop. Since we could not resolve this issue, we used fq instead of fq\_codel, for which this issue didn't occur. For fq, we set the buffer to the same size as for pfifo (1 bandwidth delay product but at least 100 packets).

>>>>

\begin{abstract}
Fair queuing is becoming increasingly prevalent in the internet and has been shown to improve performance in many circumstances. 
Performance could be improved even more if endpoints could detect the presence of fair queuing on a certain path and adjust their congestion control accordingly. 
If fair queuing is detected, the congestion control would not have to take cross traffic into account, which allows for more flexibility. 
In this paper, we develop the first algorithm that continuously checks if fair queuing is present on a path, with very high accuracy. 
When fair queuing is detected, a different congestion control can be chosen, which can result in reduced latency. Also, each flow can then specify how much queuing delay it allows, meaning that it can choose its own tradeoff between throughput and latency. 
\end{abstract}

@@

In our previous work we proposed a technique which determines the presence of fair queuing at flow startup, which works as follows: 
If fair queuing is successfully detected at flow startup, a congestion control was used which aimed to keep queuing delay low (delay-based congestion control). 
While this delay-based congestion control achieved high throughput and low delay, similar to the Vegas congestion control algorithm \cite{brakmo_tcp_1995}, 
it was vulnerable to be outcompeted by other network flows sending more aggressively, such as \cite{cardwell_bbr_2016,dong_pcc_2015,ha_cubic_2008}.
This means that our delay-based congestion control performed well but only when it wouldn't have to compete with other flows. Thus it is only used if fair queuing is detected. 
If the absence of fair queuing is detected, our previous approach uses a more aggressive congestion control (PCC \cite{dong_pcc_2015}), which can compete better with other, more aggressive network flows,
but doesn't keep delay as low as our delay-based congestion control.

@@

We would like to have an approach which 
\begin{itemize}
    \item \textbf{continuously checks} for the presence or absence of fair queuing, not only at flow startup. 
    \item \textbf{doesn't cause packet loss} while trying to determine if there is fair queuing or not. 
    \item can be \textbf{transparently used} on top of any congestion control algorithm. 
    \item lets each application \textbf{choose how much delay} it allows in case fair queuing is detected. Allowing a higher queuing delay can result in higher throughput. 
\end{itemize}

@@

On the other side, if there is no fair queuing, when the sending rate exceeds the capacity of the bottleneck link, a queue is going to build up and queuing delay is going to rise for \textbf{both} the dominant and the non-dominant subflow (\autoref{fig:illustration_no_fq}). 
In this case, Tonopah doesn't behave different from a network flow which doesn't have fair queuing detection. 
This can be a potential advantage which a Tonopah-enabled congestion control can have over other congestion control algorithms such as BBR, which has been shown to be unfair to other flows sometimes \cite{ware_modeling_2019,hock_experimental_2017}.

@@

Both subflows are controlled by the same congestion window. The differences in sending rates are achieved by using pacing. 
The basis of the implementation of Tonopah is \textit{Picoquic}\footnote{\url{https://github.com/private-octopus/picoquic/tree/master/picoquic}}. 
Picoquic includes support for Multipath QUIC and pacing, which we need for the implementation of Tonopah. Also, Picoquic implements several congestion control algorithms. 
We base Tonopah on the NewReno implementation of Picquic. When fair queuing is detected, the congestion window is reduced by $\frac{1}{8}$. 
This worked well for our proof-of-concept implementation but one could also specify any other behavior in case fair queuing is detected or even do nothing at all.

@@

We emulate the network using \textit{mininet} and use Linux's \textit{pfifo} queuing discipline in the case there is no fair queuing and \textit{fq} \textit{fq\_codel} for fair queuing. 
For the evaluation we choose a simple network in which the sender is connected to the receiver via a switch.

@@

To evaluate Tonopah's performance under crosstraffic, we use iperf3 with the NewReno congestion control. We start iperf 4\,s before Tonopah so that it has time to saturate the link. 
\autoref{fig:no_fq_crosstraffic} and \autoref{fig:fq_crosstraffic} show that also in the presence of cross-traffic Tonpah's detection accuracy is virtually unchanged or even slightly better. We tried to use fq\_codel as in \autoref{fig:fq} but we encountered problems when using fq\_codel in combination with iperf3. Frequently, iperf3 would lose all packets, have a timeout after a couple of seconds and stop. Since we could not resolve this issue, we used fq instead of fq\_codel, for which this issue didn't occur.
